Fail Faster

### 2025-03-19 paper/paper.tex
csplitmix

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
using both \bq's default \texttt{splittable\_random} and our improved SplitMix, as discussed in \secref{subsection:faster-rng}.

@@

The runtime of these generators over a range of sizes is shown in \figref{fig:time}. Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling. In this case, performance improvement is primarily attributable the enhanced SplitMix \rand.
In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators. The same holds for boolean lists, as sampling booleans is inexpensive relative to integers. Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.

These results show that staging and \rand choice serve distinct but complementary roles in generator performance. Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging. Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
\subsubsection{Scala}
To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala. We implement a subset of our strategies from \secref{subsection:benchmarking} in both staged and unstaged ScalaCheck.

As shown in \figref{fig:scala}, staged ScalaCheck achieves an even greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups. These results illustrate that staging is a powerful \textit{and} general optimization technique, delivering substantial performance gains in strict functional languages beyond OCaml.
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s optimized SplitMix. Performance benefits scale with generation size; lower is better.}
  \label{fig:time}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board. As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from faster SplitMix.}
  \label{fig:etna}
\end{figure}

@@

\section{Other Languages and Libraries}
\label{subsection:other-langs}

>>>>

\subsection{Benchmarking Generator Speed}
\label{subsection:benchmarking}
To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
using both \bq's default \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed in \secref{subsection:faster-rng}.

@@

The runtime of these generators over a range of sizes is shown in \figref{fig:time}. Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling. In this case, performance improvement is primarily attributable the \csplitmix \rand.
In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators. The same holds for boolean lists, as sampling booleans is inexpensive relative to integers. Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators. \tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}

These results show that staging and \rand choice serve distinct but complementary roles in generator performance. Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging. Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
\subsubsection{Scala}
To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala. We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq. This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}. Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
Staging allows us to eliminate this overhead, resulting in substantial speedups.

These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix. Performance benefits scale with generation size; lower is better.}
  \label{fig:time}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board. As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from \csplitmix.}
  \label{fig:etna}
\end{figure}

@@

\section{Other Languages and Libraries}
\tr{Subtitle: ``Beyond strict functional languages?''}
\label{subsection:other-langs}
